---
title: "Introducing the anyFit R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing the anyFit R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# 1. Introduction

The statistical and stochastic analysis of hydroclimatic data is a fundamental
step for the study of the underlying physical processes, as well as for the
related engineering tasks of design, operation and risk assessment. Today,
hydroclimatic data are increasingly utilised in risk-based decision making across
a wide range of scientific fields and industries. In response, the community has
released a plethora of datasets, ranging from centuries-long point measurements to
multi-decadal, high-resolution gridded re-analysis and observation products.

Hydroclimatic variables, however, exhibit a wide array of distinct statistical
features that make their study a demanding task. For example:

* Seasonality and cyclo-stationarity
* Over-annual trends and long-term persistence
* Intermittency (presence of zero values)
* Highly skewed, heavy-tailed distributions
* Different forms of auto- and cross-correlation

Consequently, tools developed for exploratory data analysis (EDA) of such data
should, at minimum, handle both point-based and gridded datasets, span multiple
temporal scales, remain extensible and scalable, and achieve computational
efficiency through parallelisation.

**anyFit** aims to address these requirements. It provides a streamlined pipeline
for performing EDA on point and gridded timeseries data, consolidating common
tasks, such as data loading, plotting, the inference of statistical measures and
distribution fitting, into a few lines of code, while admitting a broad range of
delimited and NetCDF files.

This vignette demonstrates that pipeline end to end on **real data**. It uses the
E-OBS gridded daily rainfall dataset that ships with the package, clipped to the
UK. Each step has a dedicated article that develops it further, with links
provided at the end.

```{r setup}
library(anyFit)
library(xts)
library(ggplot2)
library(patchwork)
```

# 2. The anyFit package

anyFit is built on the eXtensible Time Series (**xts**) data format, which
provides an extensible timeseries class by extending the well-known **zoo**
package. All visualisation functions are implemented using **ggplot2** and
**patchwork**, so every figure produced by anyFit is a ggplot object and inherits
the full customisability of these packages.

A key functionality is the fitting of distribution functions using the method of
**L-moments**. L-moments are summary statistics, analogous to the conventional
product moments (mean, variance, skewness, kurtosis), but based on linear
combinations of ordered data values. They are more robust to outliers, provide
more reliable estimates for the skewed and heavy-tailed distributions common in
environmental data, and, unlike Maximum Likelihood Estimation, do not rely on the
i.i.d. assumption, which is routinely violated by the autocorrelation present in
hydroclimatic data. anyFit supports sixteen distributions, from the Exponential
and Gamma to flexible multi-parameter families such as the Burr type-XII and the
Dagum. See `vignette("distribution-fitting")` for the full treatment.

A final design aspect is anyFit's native handling of **intermittent** processes,
i.e. the presence of zero values. Across all functions, including the fitting
functions, the package provides the option to exclude (or include) zeros via a
cut-off threshold (`ignore_zeros`, `zero_threshold`), and typical intermittency
measures such as the probability dry (`Pdr`) are computed by default.

# 3. Loading gridded data

anyFit is structured around the **load–process–visualize** paradigm. The package
ships an example E-OBS file, ensemble-mean daily rainfall over Europe on a
0.25° grid (2011–2022), which we locate with `system.file()`.

Gridded NetCDF data are loaded with `nc2xts()`, which reads the file directly and
returns an **`sxts`** (spatial xts) object: a single structure that carries the
timeseries together with its coordinates and projection. `nc2xts()` can clip the
data on loading using a country name, a continent or a bounding box, reading only
the requested spatial hyperslab from disk. Here we clip to the UK.

```{r}
f <- system.file("extdata", "rr_ens_mean_0.25deg_reg_2011-2022_v27.0e.nc",
                 package = "anyFit")

# world_data stores the UK under this exact name
uk_name <- "U.K. of Great Britain and Northern Ireland"

rain_uk <- nc2xts(f, varname = "rr", country = uk_name)
summary(rain_uk)
```

The result is an `sxts` with one column per grid cell and one row per day. The
`sxts` class and its spatial operations are covered in detail in
`vignette("sxts-class")`.

Because `nc_ggplot()` draws one map per timestep, we visualise a single wet winter
day as a snapshot. As the returned object is a ggplot, we overlay the UK coastline
using standard ggplot2 syntax.

```{r}
nc_ggplot(rain_uk["2015-12-05"], legend.title = "Rainfall (mm)",
          viridis.option = "mako") &
  borders("world", regions = "UK")
```

# 4. Processing: aggregation and statistics

Hydroclimatic processes act on different temporal scales, so it is useful to
aggregate the data. `period_apply_nc()` aggregates a gridded `sxts` over a chosen
temporal scale while preserving the spatial metadata. Here we compute mean annual
rainfall totals.

```{r}
annual_uk <- period_apply_nc(rain_uk, period = "years", FUN = "sum")
mean_annual <- basic_stats_nc(annual_uk)[["Mean"]]

nc_ggplot(mean_annual, legend.title = "Mean annual\nrainfall (mm)",
          viridis.option = "mako") &
  borders("world", regions = "UK") &
  theme(legend.key.width = grid::unit(1.2, "cm"))
```

With the data at hand, `basic_stats_nc()` computes a panel of statistical measures
per cell (moments, L-moments, quantiles and intermittency measures) and returns
them as a multi-layer raster, which we map with `nc_ggplot()`. Selecting layers by
name keeps the code readable.

```{r}
gstats <- basic_stats_nc(rain_uk, ignore_zeros = TRUE)
nc_ggplot(gstats[[c("Mean", "Pdr")]], viridis.option = "mako") &
  borders("world", regions = "UK")
```

# 5. Distribution fitting

anyFit uses the method of L-moments to infer marginal distribution parameters. For
gridded data this is implemented in `fitlm_nc()`, which fits and compares multiple
candidate distributions at once. Here we test two candidates of increasing
flexibility, the 2-parameter Gamma and the 3-parameter Gamma. (The dedicated
article fits more candidates and adds L-ratio diagnostics.)

```{r, fig.height = 5}
fits <- fitlm_nc(rain_uk, ignore_zeros = TRUE,
                 candidates = c("gamma", "gamma3"))
fits$gof_plots
```

`fitlm_nc()` returns, for each candidate, rasters of the fitted parameters, the
theoretical and sampled L-moments, and the goodness-of-fit metrics. For example,
the fitted 3-parameter Gamma parameters map as:

```{r, fig.height = 4}
nc_ggplot(fits$fit_results$gamma3$raster_params, viridis.option = "mako") &
  borders("world", regions = "UK") &
  theme(legend.text = element_text(size = 7))
```

# 6. Point processes

Many applications require the analysis of point-based timeseries, such as station
observations or a series extracted from a gridded product. A point series is
extracted from the same NetCDF file with `nc2xts_nn()`, which returns the value of
the nearest grid cell to the supplied coordinates.

```{r}
london <- nc2xts_nn(f, varname = "rr", coords = data.frame(x = -0.1278, y = 51.5074))
colnames(london) <- "London"
```

`basic_stats()` streamlines the essential facets of a single series (the raw
series, the empirical PDF, the empirical CDF and the autocorrelation function) in
one call, while also computing a table of summary statistics.

```{r, fig.height = 5}
bstats <- basic_stats(london, pstart = "2015", pend = "2018", plot = TRUE,
                      show_table = FALSE, show_label = FALSE, ignore_zeros = TRUE)
bstats$plot
```

The point-analysis toolkit, comprising monthly (cyclo-stationary) fitting,
autocorrelation structures and dependence diagnostics, is covered in
`vignette("point-seasonal")`.

# 7. From analysis to stochastic modelling

The outputs of anyFit are designed to feed directly into downstream stochastic
simulation. The fitted marginal distributions and autocorrelation structures
provide exactly the parameters required by simulation engines such as the
**anySim** package. The block below reproduces a univariate process with the
AutoRegressive To Anything (ARTA) model, parameterised from a 3-parameter Gamma
fit and a fitted Cauchy-type autocorrelation structure. It runs only if `anySim`
is installed.

```{r}
if (requireNamespace("anySim", quietly = TRUE)) {
  params <- fitlm_gamma3(london, ignore_zeros = TRUE)
  acfs   <- fit_ACF(london, lag_max = 10, ignore_zeros = TRUE)

  ACS <- anySim::csCAS(param = c(acfs$ACF_params$CAS$beta,
                                 acfs$ACF_params$CAS$kappa), lag = 500)
  ARTApar <- anySim::EstARTAp(ACF = ACS, dist = "qgamma3",
                              params = params$Param, NatafIntMethod = "GH")
  sim <- anySim::SimARTAp(ARTApar = ARTApar, steps = 2000)

  sim_xts <- xts(as.numeric(unlist(sim$X)),
                 order.by = seq(as.POSIXct("2100-01-01", tz = "UTC"),
                                by = "day", length.out = 2000))

  fit_diagnostics(sim_xts, dist = "gamma3", params = params$Param,
                  ignore_zeros = TRUE)$Diagnostic_Plots
}
```

# 8. Where to next

This tour touched every stage of the pipeline. Each has a dedicated article:

* `vignette("sxts-class")`: the spatial-xts class, its accessors and spatial
  operations (masking, zonal statistics, raster conversion).
* `vignette("distribution-fitting")`: L-moment fitting across a grid, goodness of
  fit, and L-ratio diagnostic diagrams.
* `vignette("point-seasonal")`: point series, monthly/cyclo-stationary fitting,
  autocorrelation structures and dependence.
* `vignette("gridded-workflow")`: the full gridded load–process–visualize
  workflow, including extreme value analysis.

anyFit offers an integrated framework for the exploratory data analysis of
hydroclimatic data. Through its `sxts` class it brings the convenience of `xts` to
the spatial domain, and through its L-moment-based fitting it provides robust
distribution estimation in the presence of autocorrelation and intermittency,
setting the foundations for a timeseries analysis and modelling ecosystem.
