---
title: "Introducing the anyFit R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing the anyFit R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# 1. Introduction

The statistical and stochastic analysis of hydroclimatic data is a fundamental
step for the study of the underlying physical processes, as well as for the
related engineering tasks of design, operation and risk assessment. Today,
hydroclimatic data are increasingly utilised in risk-based decision making across
a wide range of scientific fields and industries. In response, the community has
released a plethora of datasets, ranging from centuries-long point measurements to
multi-decadal, high-resolution gridded re-analysis and observation products.

Hydroclimatic variables, however, exhibit a wide array of distinct statistical
features that make their study a demanding task. For example:

* Seasonality and cyclo-stationarity
* Over-annual trends and long-term persistence
* Intermittency (presence of zero values)
* Highly skewed, heavy-tailed distributions
* Different forms of auto- and cross-correlation

Importantly, many of these characteristics act on different temporal scales,
ranging from very fine (e.g. hourly) to very large (e.g. annual or decadal), and
they are often spatially inhomogeneous. Consequently, tools developed for
exploratory data analysis (EDA) of such data should, at minimum:

* Handle both point-based and gridded datasets
* Handle multiple temporal scales (from seconds to years)
* Be extensible and scalable
* Achieve computational efficiency through parallelisation

**anyFit** aims to address the above requirements in a robust manner. It provides
a streamlined pipeline for performing EDA on point and gridded timeseries data,
tracking — and eventually simplifying — the workflow that an experienced
practitioner would follow. It consolidates common tasks, such as data loading,
plotting, inference of simple statistical measures and distribution fitting, into
a few lines of code, while admitting a broad range of delimited and NetCDF files
with strict and non-strict timesteps.

```{r setup}
library(anyFit)
library(xts)
```

# 2. The anyFit package

anyFit is built on the eXtensible Time Series (**xts**) data format, a powerful
package that provides an extensible timeseries class, enabling uniform handling of
many R timeseries classes by extending the well-known **zoo** package. All
visualisation functions are implemented using **ggplot2** and **patchwork**. In
this way, every figure produced by anyFit is a ggplot object and, as such,
inherits the full customisability of these packages, reducing both complexity and
the number of arguments required by the plotting functions.

A key functionality of the package is the fitting of distribution functions. Here
we have opted to focus on methods based on **L-moments**. L-moments
are summary statistics, analogous to conventional product moments (mean, variance,
skewness, kurtosis), but based on linear combinations of ordered data values. They
are regarded as more robust because they are less sensitive to outliers and provide
more reliable estimates for the skewed, heavy-tailed distributions common in
environmental data. Moreover, hydroclimatic processes exhibit autocorrelation,
which violates the i.i.d. assumption underlying Maximum Likelihood Estimation (MLE);
L-moments do not rely on this assumption and are therefore more suitable for data
where autocorrelation is present, which is almost always the case for
hydroclimatic data.

The estimation of distribution parameters by the method of L-moments is implemented
in a two-way manner. For distributions whose L-moments admit an analytic, explicit
expression (e.g. Gamma, Weibull, Gumbel, Log-Normal), the analytical equations are
used. anyFit additionally extends support to more flexible, multi-parameter
distributions that are not commonly encountered in statistical R-packages, namely
the Generalised Gamma, the Generalised Gamma with location, the Burr type-XII and
the Dagum. These distributions provide substantial flexibility in capturing heavy
tails and complex tail behaviour, but lack closed-form L-moment expressions. For
them, the package adopts a numerical optimisation procedure (the L-BFGS-B
quasi-Newton method) that minimises the error between the theoretical and sampled
L-moments, using initial values drawn from pre-computed tables that span the
feasible L-ratio parameter space of each distribution. The combination of a fast
optimiser with good initial values yields an effective and efficient fitting
procedure, which, for gridded datasets, can be further accelerated through
parallel computation.

In detail, the distributions supported by anyFit are:

* Exponential, Rayleigh, Gamma, Normal, Log-Normal
* Generalized Logistic, 3-parameter Weibull, Gumbel
* 3-parameter Gamma (Pearson III)
* Generalized Extreme Value (GEV), Generalized Pareto (GPD)
* Generalized Gamma, Generalized Gamma with location
* Burr type-XII, Dagum, Exponentiated Weibull

The distribution functions follow the standard R notation (`d`/`p`/`q`/`r` for
density, CDF, quantile and random generation, e.g. `qgamma3`, `pburr`, `rdagum`),
while the L-moment fitting functions share the `fitlm_` prefix followed by the
distribution abbreviation (e.g. `fitlm_gev`, `fitlm_burr`).

A final design aspect of anyFit is its native handling of **intermittent**
processes, i.e. the presence of zero values in a timeseries. Intermittency is a key
property of processes such as precipitation at fine scales. Across all functions,
including the fitting functions, the package provides the option to exclude (or
include) zeros via a cut-off threshold (`ignore_zeros`, `zero_threshold`), and
typical intermittency measures such as the probability dry (`Pdr`) are included in
the default statistical measures.

# 3. Package architecture and the `sxts` class

anyFit is structured around the **load–process–visualize** paradigm. The *load*
module imports timeseries data from delimited files (`delim2xts()`) and NetCDF
datasets (`nc2xts()`). The *process* module handles missing values, computes
statistical summaries, aggregates data across temporal scales, and fits candidate
distributions. The *visualize* module offers a range of plotting functions
(ridge plots, boxplots, violin plots, raster maps).

The package adopts systematic naming conventions. Loading functions follow a
`source2format` pattern (`delim2xts()`, `nc2xts()`). Functions operating on gridded
NetCDF data carry the `_nc` suffix (`basic_stats_nc()`, `fitlm_nc()`,
`period_apply_nc()`), while their point-based counterparts use `_xts` or no suffix
(`aggregate_xts()`, `basic_stats()`). Monthly-scale analysis is grouped under the
`monthly_` prefix, and autocorrelation functions use the `_ACF` suffix.

To handle gridded timeseries efficiently, anyFit introduces a dedicated object
class, the **`sxts`** (spatial xts) class. `sxts` is an S3 extension of `xts` that
stores spatial attributes, coordinates and projection, alongside the timeseries
data, with rows indexing time and columns indexing spatial locations. This enables
seamless spatial operations (subsetting, masking, zonal statistics, raster
conversion) without resorting to external raster intermediates, while inheriting
the full temporal machinery of `xts`.

An `sxts` object can be constructed directly from a data matrix, a time index, a
table of coordinates and a projection string. To keep this vignette lightweight and
fully reproducible, we work exclusively with **synthetic data**. To make the
synthetic *fields* look realistic, we define a generator that reproduces the salient
features of a precipitation field, namely:

* **spatial correlation** — neighbouring cells behave similarly, so that wet and dry
  areas form coherent patches rather than random noise. This is achieved by drawing
  the daily fields from a multivariate Gaussian whose covariance decays
  exponentially with distance (via a Cholesky factor);
* an **orographic mean field** — the wet-day mean intensity varies smoothly in
  space (here, wetter to the north plus a localised ridge), mimicking the influence
  of topography;
* **seasonality** — the probability of a wet day varies with the calendar month
  (wetter in winter);
* **intermittency** — dry days are produced by thresholding the latent field.

```{r}
# Spatially correlated, seasonal, intermittent precipitation field.
# Returns a [n_locations x n_time] matrix (rows = cells, columns = days).
sim_precip_field <- function(coords, dates, seed = 1, cor_range = 0.8, shape = 1.3) {
  set.seed(seed)
  n <- nrow(coords); nt <- length(dates)

  # Exponential spatial covariance -> correlated daily Gaussian fields
  D <- as.matrix(dist(coords[, c("x", "y")]))
  L <- chol(exp(-D / cor_range) + diag(1e-6, n))
  Z <- crossprod(L, matrix(rnorm(n * nt), n, nt))         # n x nt, spatially correlated

  # Normalised coordinates, used to drive the spatial gradients
  gx <- (coords$x - min(coords$x)) / diff(range(coords$x))
  gy <- (coords$y - min(coords$y)) / diff(range(coords$y))

  # Wet-day probability: seasonal (wetter in winter) and spatial (wetter to the
  # north) -> per cell-day occurrence threshold on the latent field
  mth  <- as.integer(format(dates, "%m"))
  pwet <- 0.25 + 0.30 * (0.5 + 0.5 * cos(2 * pi * (mth - 1) / 12))   # per day
  swet <- 0.7 + 0.6 * gy                                            # per cell
  wet  <- Z > qnorm(1 - pmin(0.9, outer(swet, pwet)))

  # Orographic wet-day mean field: wetter to the north plus a smooth ridge
  mu <- 2 + 5 * gy + 4 * exp(-((gx - 0.3)^2 + (gy - 0.7)^2) / 0.08)

  # Gamma wet-day intensity, driven by the same field (intense days are wetter)
  amount <- qgamma(pnorm(Z), shape = shape, scale = mu / shape)
  wet * amount
}
```

We use it to build a tiny `sxts` on a regular grid placed over Belgium (so that the
country-based spatial operations shown later are meaningful):

```{r}
dates_s  <- seq(as.Date("2018-01-01"), by = "day", length.out = 365 * 2)
coords_s <- expand.grid(x = c(4.0, 4.5, 5.0), y = c(50.0, 50.5))
mat_s    <- sim_precip_field(coords_s, dates_s, seed = 7)   # [cells x days]

grid_demo <- sxts(t(mat_s), order.by = as.POSIXct(dates_s, tz = "UTC"),
                  coords = coords_s, projection = "+proj=longlat +datum=WGS84")

summary(grid_demo)
```

The class ships with accessors for its spatial attributes and an `[` method that is
aware of the spatial dimension: subsetting by time (rows) preserves all locations,
whereas subsetting by location (columns) updates the stored coordinates and the
element count accordingly.

```{r}
coords(grid_demo)        # spatial coordinates
projection(grid_demo)    # coordinate reference system
elements(grid_demo)      # number of spatial locations

# Spatial subset: keep the first three locations
sub <- grid_demo[, 1:3]
elements(sub)
```

# 4. Gridded workflow on synthetic data

In practice, gridded data are most often stored in NetCDF files. We therefore write
a small synthetic NetCDF and demonstrate the full gridded pipeline on it, exactly as
one would with a real product. Using the field generator defined above, we produce
six years of daily precipitation on a 7 × 5 grid over Belgium.

```{r}
lons    <- seq(3.0, 6.0, by = 0.5)    # 7 longitudes
lats    <- seq(49.7, 51.3, by = 0.4)  # 5 latitudes
dates_g <- seq(as.Date("2015-01-01"), by = "day", length.out = 365 * 6)

# Generate the field. Cell rows follow expand.grid(x = lons, y = lats), i.e. the
# lon-fastest column order that nc2xts() rebuilds, so no further reordering is needed.
coords_g <- expand.grid(x = lons, y = lats)
cell_mat <- sim_precip_field(coords_g, dates_g, seed = 1)         # [cells x days]
arr      <- array(cell_mat, dim = c(length(lons), length(lats), length(dates_g)))

nc_file <- tempfile(fileext = ".nc")
londim  <- ncdf4::ncdim_def("lon", "degrees_east",  lons, longname = "longitude")
latdim  <- ncdf4::ncdim_def("lat", "degrees_north", lats, longname = "latitude")
timedim <- ncdf4::ncdim_def("time", "days since 2015-01-01 00:00:00",
                            0:(length(dates_g) - 1), unlim = TRUE, longname = "time")
vardef  <- ncdf4::ncvar_def("rr", "mm", list(londim, latdim, timedim), -9999,
                            longname = "precipitation")

nc <- ncdf4::nc_create(nc_file, vardef)
ncdf4::ncvar_put(nc, vardef, arr)
ncdf4::nc_close(nc)
```

The data are loaded with `nc2xts()`, which returns an `sxts` object. Since our
synthetic file carries no CRS metadata, we supply the projection explicitly. The
same function can conveniently clip the data on loading, using a country name, a
continent or a user-supplied shapefile.

```{r}
rain_grid <- nc2xts(filename = nc_file, varname = "rr",
                    projection = "+proj=longlat +datum=WGS84")
rain_grid
```

The resulting `sxts` carries one column per grid cell and one row per day. Because
`nc_ggplot()` draws one map per timestep, we visualise a single day as a snapshot.
As `nc_ggplot()` is built on ggplot2 and patchwork, the returned object can be
freely customised afterwards.

```{r}
nc_ggplot(rain_grid[150], legend.title = "Precip (mm)", viridis.option = "turbo")
```

Hydroclimatic processes act on different temporal scales, so it is useful to
aggregate the data. `period_apply_nc()` aggregates a gridded `sxts` over a chosen
temporal scale, preserving the spatial metadata. Here we aggregate to annual totals.

```{r}
annual_grid <- period_apply_nc(rain_grid, period = "years", FUN = "sum")
nc_ggplot(annual_grid, title = TRUE, viridis.option = "turbo",
          legend.title = "Annual precip (mm)", common_legend = TRUE)
```

With the data at hand, we infer basic statistical measures across the grid. The
`basic_stats_nc()` function computes 34 statistical measures per cell — including
moments, L-moments, quantiles and intermittency measures — and returns them as a
multi-layer raster, which we map with `nc_ggplot()`. Selecting layers by name keeps
the code readable.

```{r}
gstats <- basic_stats_nc(rain_grid, ignore_zeros = TRUE)
nc_ggplot(gstats[[c("Mean", "Pdr")]], viridis.option = "turbo")
```

## 4.1 Marginal distribution fitting

anyFit uses the method of L-moments to infer marginal distribution parameters. For
gridded data this is implemented in `fitlm_nc()`, which fits and compares multiple
candidate distributions at once. We test three candidates of increasing
flexibility: the 2-parameter Gamma, the 3-parameter Gamma, and the flexible Burr
type-XII.

```{r}
candidates <- c("gamma", "gamma3", "burr")
fits_all   <- fitlm_nc(rain_grid, ignore_zeros = TRUE, candidates = candidates)
```

To assess goodness of fit, `fitlm_nc()` outputs a density plot of the implemented
goodness-of-fit measures for every candidate distribution. These measures include
the negative log-likelihood (MLE), the Cramér-von Mises (CM) and Kolmogorov-Smirnov
(KS) statistics, the mean squared error of quantiles (MSEquant), the percentage
difference of maxima (DiffOfMax) and the mean difference of the ten largest values
(MeanDiffOf10Max). The density merely summarises the range of each measure across
all grid cells, providing a visual aid for determining which distribution behaves
best for most of the data.

```{r, fig.height = 5}
fits_all$gof_plots
```

For a deeper analysis, the function returns rasters with the fitted parameters,
the theoretical and sampled L-moments, and the goodness-of-fit metrics for each
distribution. For example, the spatial distribution of the fitted Burr type-XII
parameters is mapped below.

```{r, fig.height = 5}
nc_ggplot(fits_all$fit_results$burr$raster_params, viridis.option = "turbo")
```

## 4.2 L-ratio plots

An informative way of assessing whether a distribution is a suitable candidate is
through L-ratio plots. Statistical distributions possess a theoretical L-ratio
support space; by comparing this space to the sampled L-ratios we can determine
whether a candidate may provide a good fit. A distribution whose theoretical support
covers the majority of the sampled L-ratios constitutes a promising candidate. This
is implemented in `LRatio_check()`, currently available for the Dagum, Generalized
Gamma, Exponentiated Weibull and Burr type-XII distributions.

```{r, fig.height = 6}
ts_lmoms <- lmom_stats(rain_grid, ignore_zeros = TRUE)
lcheck   <- LRatio_check(ts_lmoms)
lcheck$multi_plots
```

## 4.3 Extreme value analysis

Extreme value analysis is routinely used to infer design levels for civil and
environmental engineering structures. The annual maxima of daily precipitation can
be derived with `period_apply_nc()` using the maximum aggregation operator, and then
fitted with the Generalised Extreme Value (GEV) distribution via `fitlm_nc()`.

```{r, fig.height = 5}
amax_grid <- period_apply_nc(rain_grid, period = "years", FUN = "max")
gev_fit   <- fitlm_nc(amax_grid, candidates = "gev")
nc_ggplot(gev_fit$fit_results$gev$raster_params, viridis.option = "turbo")
```

# 5. Spatial operations on `sxts`

Beyond statistical analysis, the `sxts` class supports a range of spatial
operations. The data can be masked by a bounding box, a shapefile, a country or a
continent. Note that the `sxts` object and the mask must share the same coordinate
reference system.

```{r}
# Bounding-box mask
bbox_grid <- mask.sxts(rain_grid, xlim = c(3.5, 4.5), ylim = c(50, 51))
elements(bbox_grid)

# Country mask
be_grid <- mask.sxts(rain_grid, mask = world_data[world_data$name == "Belgium", ])
elements(be_grid)
```

`zonal_stats()` aggregates the point values of an `sxts` over polygon boundaries —
supplied as a shapefile, a country or a continent — and returns a plain `xts` with
one column per polygon. Below we compute the spatial mean precipitation over
Belgium for each timestep.

```{r}
be_mean <- zonal_stats(rain_grid, country = "Belgium", FUN = mean)
head(be_mean)
```

# 6. Point processes

While the preceding sections focused on gridded datasets, many practical
applications require the analysis of point-based timeseries, such as station
observations or series extracted from gridded products. anyFit applies the same
unified statistical methodology to point data.

A point timeseries can be extracted from a gridded NetCDF file using `nc2xts_nn()`,
which returns the value of the nearest grid cell to the supplied coordinates (no
spatial interpolation is performed).

```{r}
point_coords <- data.frame(x = 4.4, y = 50.4)
point_xts    <- nc2xts_nn(filename = nc_file, varname = "rr", coords = point_coords)
```

The `basic_stats()` function streamlines the production of the essential facets of a
timeseries — the raw series, the empirical PDF, the empirical CDF and the
autocorrelation function (ACF) — in a single call, while also computing a table of
summary statistical measures.

```{r}
bstats <- basic_stats(point_xts, pstart = "2016", pend = "2018",
                      show_table = FALSE, show_label = FALSE, ignore_zeros = TRUE)
bstats$plot
```

## 6.1 Distribution fitting and diagnostics

For a single series, `fitlm_multi()` fits a list of candidate distributions and
returns the fitted parameters, a goodness-of-fit summary, and Q-Q / P-P diagnostic
plots. Q-Q and P-P plots compare theoretical and empirical quantiles and
probabilities respectively; a perfect fit falls on the 1:1 line.

```{r, fig.height = 5}
fits <- fitlm_multi(point_xts, candidates = list("gamma", "gamma3", "burr"),
                    ignore_zeros = TRUE)
fits$diagnostics
```

```{r}
fits$GoF_summary
```

## 6.2 Monthly (cyclo-stationary) analysis

A notable feature of anyFit is the fitting of candidate distributions per month,
which supports the study of seasonality and cyclo-stationary processes. We first
aggregate the series to a monthly scale, then fit distributions per month and
inspect the monthly Q-Q plots.

```{r, fig.height = 9, fig.width = 8}
agg_ts      <- aggregate_xts(ts = point_xts, periods = c("months", "years"), FUN = "sum")
smonthly_ts <- agg_ts$list_months$aggregated

monthly_fits <- fitlm_monthly(smonthly_ts, candidates = list("exp", "gamma", "gamma3"),
                              ignore_zeros = TRUE)
monthly_fits$monthly_QQplot
```

Visualisation plays an essential role in data comprehension. anyFit provides
wrappers for boxplots and violin plots that group a variable by month, offering an
intuitive view of inter-annual variability.

```{r}
monthly_boxplots(point_xts, palette = "Set3", ignore_zeros = TRUE)
```

```{r}
monthly_violins(point_xts, palette = "Set3", ignore_zeros = TRUE)
```

When studying cyclo-stationarity it is also important to account for
autocorrelation. anyFit fits three theoretical autocorrelation structures — the
Cauchy (CAS), the Hurst-Kolmogorov (HK) and the exponential Short-Range Dependence
(SRD) — both globally with `fit_ACF()` and on a monthly basis with
`fit_ACF_monthly()`.

```{r}
acfs <- fit_ACF(point_xts, lag = 10, ignore_zeros = TRUE)
acfs$ACF_plot
```

## 6.3 Dependence between processes

Dependencies between different stations or processes are often of interest. Moving
beyond the traditional correlation coefficient, `correl_plots()` plots the joint
distribution of two timeseries in three spaces: the original space (scatter plot),
the uniform space (copula) and the standard normal space. The copula plot depicts
the full dependence structure without the influence of the marginal distributions,
and is particularly useful for examining asymmetries and tail dependencies.

```{r, fig.height = 4}
# A second series extracted from a neighbouring grid cell: thanks to the spatial
# correlation built into the field, the two series are naturally dependent.
point2  <- nc2xts_nn(filename = nc_file, varname = "rr",
                     coords = data.frame(x = 5.1, y = 50.9))

correls <- correl_plots(point_xts, point2, check_common = TRUE, ignore_zeros = TRUE)
correls$combined
```

# 7. From analysis to stochastic modelling

The outputs of anyFit are designed to feed directly into downstream stochastic
simulation. The fitted marginal distributions and autocorrelation structures
provide exactly the parameters required by simulation engines such as the
**anySim** package, enabling Monte-Carlo risk assessment and scenario generation.

The following example reproduces a univariate process with the AutoRegressive To
Anything (ARTA) model, parameterised entirely from the quantities we inferred above
(the fitted distribution and the fitted Cauchy autocorrelation structure). The
block runs only if `anySim` is installed.

```{r}
if (requireNamespace("anySim", quietly = TRUE)) {
  params <- fitlm_gamma3(point_xts, ignore_zeros = TRUE)

  # Target autocorrelation structure (Cauchy), from fit_ACF above
  ACS <- anySim::csCAS(param = c(acfs$ACF_params$CAS$beta,
                                 acfs$ACF_params$CAS$kappa), lag = 500)

  ARTApar <- anySim::EstARTAp(ACF = ACS, dist = "qgamma3",
                              params = params$Param, NatafIntMethod = "GH")
  sim     <- anySim::SimARTAp(ARTApar = ARTApar, steps = 2000)

  sim_xts <- xts(as.numeric(unlist(sim$X)),
                 order.by = seq(as.POSIXct("2100-01-01", tz = "UTC"),
                                by = "day", length.out = 2000))

  fit_diagnostics(sim_xts, dist = "gamma3", params = params$Param,
                  ignore_zeros = TRUE)$Diagnostic_Plots
}
```

Similarly, the month-to-month statistics from `monthly_stats()` together with the
monthly distribution fits support the simulation of a cyclo-stationary monthly
process via the Stochastic Periodic ARTA (SPARTA) model.

```{r}
if (requireNamespace("anySim", quietly = TRUE)) {
  monthly_sts <- monthly_stats(point_xts, aggregated = TRUE, FUN = "sum")

  NumOfSeasons <- 12
  rtarget <- monthly_sts$lag1$Value
  rtarget[is.na(rtarget)] <- 0.1
  FXs <- rep("qzi", NumOfSeasons)

  PFXs <- vector("list", NumOfSeasons)
  for (i in 1:NumOfSeasons) {
    PFXs[[i]] <- c(p0 = monthly_sts$agg_stats["Pdr", i],
                   Distr = match.fun("qgamma3"),
                   setNames(monthly_fits$params_monthly$gamma3[[i]],
                            rownames(monthly_fits$params_monthly$gamma3[i])))
  }

  SPARTApar <- anySim::EstSPARTA(s2srtarget = rtarget, dist = FXs, params = PFXs,
                                 NatafIntMethod = "GH", NoEval = 9,
                                 polydeg = 8, nodes = 11)
  simS <- anySim::SimSPARTA(SPARTApar = SPARTApar, steps = 200)

  str(simS$X)
}
```

# 8. Conclusion

anyFit offers an integrated and efficient framework for the exploratory data
analysis of hydroclimatic data, tailored to modern high-resolution point and
gridded datasets. Through its `sxts` class it brings the convenience of `xts` to
the spatial domain, and through its L-moment-based fitting it provides robust
distribution estimation in the presence of autocorrelation and intermittency. While
designed with hydroclimatic applications in mind, its data-agnostic core makes it
applicable wherever timeseries exhibit seasonality, autocorrelation, intermittency
or skewed distributions. Its outputs integrate naturally with downstream stochastic
simulation packages such as anySim, setting the foundations for a timeseries
analysis and modelling ecosystem.
