---
title: "Gridded workflow and visualization"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(anyFit)
library(xts)
library(ggplot2)
library(patchwork)
```

# 1. The gridded pipeline

anyFit follows a **load → process → visualize** paradigm. This article follows the
gridded branch of that pipeline end to end on the bundled E-OBS daily rainfall
file, and closes with an extreme-value analysis. Throughout, the data reside in a
single `sxts` object (see `vignette("sxts-class")`).

```{r}
f <- system.file("extdata", "rr_ens_mean_0.25deg_reg_2011-2022_v27.0e.nc",
                 package = "anyFit")
uk_name <- "U.K. of Great Britain and Northern Ireland"
```

# 2. Load: reading and clipping NetCDF

`nc2xts()` reads a NetCDF variable directly and returns an `sxts`. It can clip the
data **while reading**, reading only the requested spatial hyperslab from disk,
via any of four options:

* `country =`: a country name from `world_data$name`;
* `continent =`: a continent name (e.g. `"Europe"`);
* `xlim =` / `ylim =`: a bounding box;
* `shapefile =`: a path to a shapefile.

Here we read a bounding box over the British Isles and map a single wet winter day.
Because the result is a ggplot, we overlay the coastline with standard ggplot2
syntax.

```{r}
brit <- nc2xts(f, varname = "rr", xlim = c(-11, 2), ylim = c(49.5, 61))

nc_ggplot(brit["2015-12-05"], legend.title = "Rainfall (mm)",
          viridis.option = "mako") &
  borders("world", regions = c("UK", "Ireland"))
```

For the analysis that follows we clip to the UK by name.

```{r}
rain_uk <- nc2xts(f, varname = "rr", country = uk_name)
```

# 3. Process: temporal aggregation

`period_apply_nc()` aggregates a gridded `sxts` over a chosen temporal scale while
preserving the spatial metadata. The aggregation function is user-specified:
`"sum"` for rainfall totals, `"mean"` for a variable such as temperature, or
`"max"` to extract block extremes. Here we form annual totals.

```{r}
annual_uk <- period_apply_nc(rain_uk, period = "years", FUN = "sum")
c(years = nrow(annual_uk), cells = elements(annual_uk))
```

# 4. Process: gridded statistics

`basic_stats_nc()` computes a panel of summary statistics at every cell (moments,
L-moments, quantiles and intermittency measures) and returns a multi-layer
raster. We map mean wet-day rainfall alongside the probability dry.

```{r, fig.height = 4}
gstats <- basic_stats_nc(rain_uk, ignore_zeros = TRUE)
nc_ggplot(gstats[[c("Mean", "Pdr")]], viridis.option = "mako") &
  borders("world", regions = "UK")
```

For seasonal structure, `monthly_stats_nc()` computes the same statistics
**separately for each calendar month**. Contrasting January with July shows the
seasonal variation in mean wet-day rainfall.

```{r, fig.height = 4}
mstats <- monthly_stats_nc(rain_uk, ignore_zeros = TRUE)

p_jan <- nc_ggplot(mstats$January[["Mean"]], legend.title = "January",
                   viridis.option = "mako")
p_jul <- nc_ggplot(mstats$July[["Mean"]], legend.title = "July",
                   viridis.option = "mako")
(p_jan | p_jul) & borders("world", regions = "UK")
```

# 5. Visualize: customising `nc_ggplot()`

Every map is a native ggplot object, so it can be customised with the full ggplot2
and patchwork vocabulary. `nc_ggplot()` itself provides several conveniences: a
`viridis.option` for the colour palette, a `legend.title`, per-panel titles via
`title = TRUE`, and `common_legend = TRUE` to share one colour scale across a
multi-layer raster. Below we map the L-scale, L-skewness and probability dry with
one shared legend.

```{r, fig.height = 4}
nc_ggplot(gstats[[c("LScale", "LSkewness", "Pdr")]],
          viridis.option = "mako", common_legend = TRUE) &
  borders("world", regions = "UK")
```

# 6. Extreme value analysis

Extreme value analysis underpins infrastructure design and flood risk assessment.
The procedure is concise: reduce the daily series to annual maxima with
`period_apply_nc(FUN = "max")`, then fit the Generalised Extreme Value (GEV)
distribution at every cell with `fitlm_nc()`.

```{r, fig.height = 4}
amax <- period_apply_nc(rain_uk, period = "years", FUN = "max")
gev  <- fitlm_nc(amax, candidates = "gev", ignore_zeros = TRUE)

nc_ggplot(gev$fit_results$gev$raster_params, viridis.option = "mako") &
  borders("world", regions = "UK")
```

The three panels are the GEV location, scale and shape parameters. The location
and scale describe the magnitude of extreme daily rainfall, while the shape
parameter governs the heaviness of the upper tail, the key control on how quickly
design levels grow with return period.

# 7. See also

* `vignette("distribution-fitting")`: the L-moment fitting engine, goodness of fit
  and L-ratio diagnostics behind `fitlm_nc()`.
* `vignette("sxts-class")`: the spatial-xts object that carries every grid through
  this pipeline.
