---
title: "Point and seasonal analysis"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(anyFit)
library(xts)
library(ggplot2)
library(patchwork)
```

# 1. Point processes

Many applications require the analysis of point-based timeseries, such as a rain
gauge, a station record, or a series extracted from a gridded product. anyFit
applies the same statistical machinery to point data as it does to grids.
Crucially, hydroclimatic point series are usually **cyclo-stationary**: both their
marginal distribution and their autocorrelation structure change with the season.
This article demonstrates the point-analysis toolkit on a single location.

We extract the London grid cell from the bundled E-OBS daily rainfall file with
`nc2xts_nn()`, which returns the nearest cell to the supplied coordinates.

```{r}
f <- system.file("extdata", "rr_ens_mean_0.25deg_reg_2011-2022_v27.0e.nc",
                 package = "anyFit")
london <- nc2xts_nn(f, varname = "rr", coords = data.frame(x = -0.1278, y = 51.5074))
colnames(london) <- "London"
```

# 2. A first look with `basic_stats()`

`basic_stats()` streamlines the essential facets of a series into one call: the
raw series, the empirical PDF, the empirical CDF, and the autocorrelation function
(ACF), alongside a table of summary statistics. We restrict the plot to a few
years for legibility, excluding dry days.

```{r, fig.height = 5}
bstats <- basic_stats(london, pstart = "2015", pend = "2018", plot = TRUE,
                      show_table = FALSE, show_label = FALSE, ignore_zeros = TRUE)
bstats$plot
```

# 3. Seasonality at a glance

Grouping the daily values by calendar month reveals the seasonal cycle. anyFit
provides ggplot wrappers for month-grouped boxplots, violins and density ridges.

```{r}
monthly_boxplots(london, palette = "Set3", ignore_zeros = TRUE)
```

```{r}
monthly_violins(london, palette = "Set3", ignore_zeros = TRUE)
```

```{r, fig.height = 5.5}
ridge_plots(london, palette = "Set3", ignore_zeros = TRUE)$plot_monthly
```

# 4. Monthly (cyclo-stationary) distribution fitting

Rather than fit one distribution to the whole series, `fitlm_monthly()` fits a set
of candidates **separately for each calendar month**, capturing how the marginal
distribution shifts through the year. The returned monthly Q–Q plot shows how well
each month's fit matches the data, where a good fit falls on the 1:1 line.

```{r, fig.height = 9, fig.width = 8}
monthly_fits <- fitlm_monthly(london, candidates = list("exp", "weibull", "gamma3"),
                              ignore_zeros = TRUE)
monthly_fits$monthly_QQplot
```

# 5. Autocorrelation structure

Cyclo-stationarity is also present in the temporal dependence. anyFit fits three
theoretical autocorrelation structures to the sample ACF, namely the Cauchy-type
(CAS), the Hurst–Kolmogorov long-range structure (HK), and the exponential
short-range dependence structure (SRD), which encode different assumptions about
how quickly memory decays. `fit_ACF()` performs this fit globally:

```{r}
acfs <- fit_ACF(london, lag_max = 12, ignore_zeros = TRUE)
acfs$ACF_plot
```

`fit_ACF_monthly()` repeats the procedure per calendar month, revealing seasonal
changes in persistence.

```{r, fig.height = 9, fig.width = 8}
monthly_acfs <- fit_ACF_monthly(london, lag = 10)
monthly_acfs$ACF_monthly_plot
```

# 6. Dependence between locations

Dependence between two series is often of interest, whether between stations or
between a site and its neighbour. Moving beyond the plain correlation coefficient,
`correl_plots()` shows the joint behaviour in three spaces: the original space
(scatter), the uniform space (copula, which isolates the dependence structure from
the marginals), and the standard-normal space.

```{r, fig.height = 4}
neighbour <- nc2xts_nn(f, varname = "rr", coords = data.frame(x = -1.9, y = 52.5))
colnames(neighbour) <- "Birmingham"

correls <- correl_plots(london, neighbour, check_common = TRUE, ignore_zeros = TRUE)
correls$combined
```

# 7. From analysis to stochastic simulation

The monthly marginal fits and the month-to-month statistics are exactly the inputs
required by the Stochastic Periodic ARTA (SPARTA) model in the **anySim** package,
enabling simulation of a cyclo-stationary monthly process. The block runs only if
`anySim` is installed.

```{r}
if (requireNamespace("anySim", quietly = TRUE)) {
  smonthly <- aggregate_xts(london, periods = "months",
                            FUN = "sum")$list_months$aggregated
  m_fits <- fitlm_monthly(smonthly, candidates = list("gamma3"), ignore_zeros = TRUE)
  m_sts  <- monthly_stats(london, aggregated = TRUE, FUN = "sum")

  NumOfSeasons <- 12
  rtarget <- m_sts$lag1$Value
  rtarget[is.na(rtarget)] <- 0.1
  FXs <- rep("qzi", NumOfSeasons)

  PFXs <- vector("list", NumOfSeasons)
  for (i in seq_len(NumOfSeasons)) {
    PFXs[[i]] <- c(p0 = m_sts$agg_stats["Pdr", i],
                   Distr = match.fun("qgamma3"),
                   setNames(m_fits$params_monthly$gamma3[[i]],
                            rownames(m_fits$params_monthly$gamma3[i])))
  }

  SPARTApar <- anySim::EstSPARTA(s2srtarget = rtarget, dist = FXs, params = PFXs,
                                 NatafIntMethod = "GH", NoEval = 9,
                                 polydeg = 8, nodes = 11)
  simS <- anySim::SimSPARTA(SPARTApar = SPARTApar, steps = 200)
  str(simS$X)
}
```

# 8. See also

* `vignette("distribution-fitting")`: the L-moment fitting engine and goodness-of-fit
  diagnostics in depth.
* `vignette("gridded-workflow")`: the same seasonal ideas applied across a grid
  with `monthly_stats_nc()` and `fitlm_monthly_nc()`.
