---
title: "The sxts (spatial xts) class"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(anyFit)
library(xts)
library(ggplot2)
library(patchwork)
```

# 1. What is an `sxts`?

Gridded hydroclimatic data are two things at once: a **time series** (many time
steps) and a **spatial field** (many locations). Workflows that maintain these
separately, a `raster` for space and an `xts` for time, incur both a memory
overhead and the constant risk of the two representations drifting out of
alignment.

anyFit's answer is the **`sxts`** (spatial xts) class. It is a lightweight S3
extension of `xts`: the data matrix has one **row per time step** and one
**column per grid cell**, and three pieces of spatial metadata are carried as
attributes:

* `coords`: a data frame of `x`/`y` coordinates, one row per column of data;
* `projection`: a CRS string describing the coordinate reference system;
* `elements`: the number of spatial locations (i.e. data columns).

Because `sxts` *extends* `xts` rather than replacing it, the entire xts toolbox
(temporal subsetting, period aggregation, date arithmetic) remains available,
while the spatial context is retained through every operation.

# 2. Constructing an `sxts`

The usual way to obtain one is to read a NetCDF file with `nc2xts()`, which
returns an `sxts` directly. We use the bundled E-OBS daily rainfall file, clipped
to the UK.

```{r}
f <- system.file("extdata", "rr_ens_mean_0.25deg_reg_2011-2022_v27.0e.nc",
                 package = "anyFit")
uk_name <- "U.K. of Great Britain and Northern Ireland"

rain_uk <- nc2xts(f, varname = "rr", country = uk_name)
```

The dedicated `str()` and `summary()` methods report the spatial and temporal
extent concisely:

```{r}
str(rain_uk)
```

You can also build an `sxts` from scratch with the `sxts()` constructor, passing a
data matrix, a time index, a coordinate table and a projection, which is useful
when the data do not come from NetCDF. The constructor validates that the number
of coordinate rows matches the number of data columns.

```{r}
set.seed(1)
dates  <- seq(as.POSIXct("2020-01-01", tz = "UTC"), by = "day", length.out = 30)
coords <- data.frame(x = c(-2, -1, 0), y = c(51, 52, 53))
vals   <- matrix(rgamma(30 * 3, shape = 1, scale = 3), nrow = 30, ncol = 3)

toy <- sxts(vals, order.by = dates, coords = coords,
            projection = "+proj=longlat +datum=WGS84")
is.sxts(toy)
```

# 3. Accessing spatial metadata

Three accessor functions expose the spatial attributes:

```{r}
elements(rain_uk)          # number of grid cells
projection(rain_uk)        # coordinate reference system
head(coords(rain_uk))      # x / y coordinates, one row per cell
```

# 4. Subsetting is spatially aware

The `[` method treats the two dimensions differently. Subsetting by **time**
(rows), using the full range of xts date-string syntax, retains every location:

```{r}
winter15 <- rain_uk["2015-12"]     # December 2015, all cells
c(rows = nrow(winter15), cells = elements(winter15))
```

Subsetting by **location** (columns) updates the stored coordinates and the
element count to match, so that the metadata remains consistent with the data:

```{r}
first_ten <- rain_uk[, 1:10]
c(cells = elements(first_ten), coord_rows = nrow(coords(first_ten)))
```

# 5. Arithmetic, lagging and differencing preserve the class

Element-wise arithmetic, comparison, `lag()` and `diff()` all return an `sxts`
with the spatial attributes preserved automatically:

```{r}
rain_cm <- rain_uk / 10          # mm -> cm
is.sxts(rain_cm)

lagged <- lag(rain_uk, 1)        # one-day lag, still an sxts
is.sxts(lagged)
```

# 6. Spatial operations native to `sxts`

## Masking

`mask.sxts()` restricts an `sxts` to a region, given as a bounding box, a
shapefile, or an `sf`/`Spatial*` object. Here we crop to a box over central
England and confirm that the cell count is reduced.

```{r}
box <- mask.sxts(rain_uk, xlim = c(-3, 0), ylim = c(51.5, 53.5))
elements(box)

nc_ggplot(basic_stats_nc(box)[["Mean"]], legend.title = "Mean (mm)",
          viridis.option = "mako") +
  borders("world", regions = "UK")
```

## Zonal statistics

`zonal_stats()` collapses the spatial dimension by aggregating all cells that fall
within one or more polygons (supplied as a shapefile, a country or a continent)
and returns a plain `xts` with one column per polygon. Below we compute the daily
spatial-mean rainfall over the UK, then aggregate it to monthly totals for a
readable series. Because the E-OBS grid contains some missing cells, we supply an
`na.rm = TRUE` mean, so that a single missing cell does not propagate to the
entire day.

```{r}
uk_mean_daily <- zonal_stats(rain_uk, country = uk_name,
                             FUN = function(x) mean(x, na.rm = TRUE))
uk_monthly <- aggregate_xts(uk_mean_daily, periods = "months",
                            FUN = "sum")$list_months$aggregated

autoplot(uk_monthly) +
  labs(y = "UK mean rainfall (mm/month)", x = NULL) +
  theme_bw()
```

## Conversion to and from rasters

When data must be passed to a raster-based tool, `rasterFromSxts()` produces a
`RasterBrick` with time as the z-dimension, and `sxtsFromRaster()` performs the
inverse conversion. The round trip is lossless for the data and reconstructs the
spatial metadata.

```{r}
brick_uk <- rasterFromSxts(rain_uk)
class(brick_uk)

back <- sxtsFromRaster(brick_uk)
c(is_sxts = is.sxts(back), cells = elements(back))
```

# 7. Concatenating along time

Reading a multi-file dataset produces a *list* of `sxts` objects that share a
grid. `rbindlist.sxts()` concatenates them into one continuous series in a single
pass, which is considerably faster than repeated pairwise `rbind`. To illustrate,
we split the UK series into two by time and recombine it.

```{r}
part1 <- rain_uk["2011/2016"]
part2 <- rain_uk["2017/2022"]

recombined <- rbindlist.sxts(list(part1, part2))
c(rows = nrow(recombined), cells = elements(recombined))
```

# 8. Benefits of using the sxts class

By carrying space and time in a single validated object, `sxts` eliminates the
risk of misalignment between the spatial and temporal components, so that the
remaining anyFit functionality, namely statistics (`basic_stats_nc()`),
aggregation (`period_apply_nc()`), fitting (`fitlm_nc()`) and plotting
(`nc_ggplot()`), operates on a single, coherent structure. See
`vignette("gridded-workflow")` for that pipeline end to end.
